BOLD Parameter Estimation Using Particle Filters

### 2010-08-16 doc/thesis/methods.tex
saw a typo and had to fix

\subsection{Choosing $P(y_k | x_k)$}
\label{sec:Methods Weighting Function}
The choice of $P(y_k | x_k)$ is the second most important design decision, behind 
the prior. While the conventional choice for an unknown distribution is the 
Gaussian, there are reasons why it may not be the best in this case.  
As noted in \autoref{sec:Introduction Noise}, the noise is not strictly Gaussian,
nor is it strictly Wiener. As with any unknown noise however, it is necessary 
to make some assumption. If the weighting function ($P(y_k | x_k)$) exactly
matches the measurement error, then the ideal particle filter will result.
Particles with $x_k$'s that repeatedly estimate $y_k$ with large residual 
will quickly have weights near 0. Thus, a weighting function that
exactly matches $P(Y(t) | X(t))$ will easily, and correctly throw out incorrect
particles.  The cost of choosing an overly broad distribution for this
function is slow convergence.  On the other hand, an overly thin 
distribution will lead to particle deprivation (all particles
being zero-weighted). Three weighting functions were 
tested. In addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have wide tales. 
Wider tailed distributions don't down-weight
particles as fast; and thus converge more slowly. 
The Laplace distribution also has the
benefit of a non-zero slope at the origin; meaning it 
differentiates between particles even near the origin.

After trial and error, for this work the Gaussian with a standard deviation 
of $0.005$ was chosen. Although using an adaptive weighting function
might be better, in tests this often led to unpredictable results. 
With some work it may be possible to set the standard deviation by
taking a small sample from resting data and using the sample standard deviation.
In this work, however, I set the standard deviation manually at $0.005$,
because it gave the best consistency.

@@

Because the prior is initially represented with significantly more
particles, if for some reason the effective
number of particles stays high, resampling could take a long time to occur.
For this reason, rather than allowing the particle filter to continue on 
with this large number of particles, after 20 seconds have passed the
algorithm forces resampling. The choice of 20 seconds is 
arbitrary, but at the very least it the particles some time to be reweighted.

>>>>

\subsection{Choosing $P(y_k | x_k)$}
\label{sec:Methods Weighting Function}
The choice of $P(y_k | x_k)$ is the second most important design decision, behind 
the prior. While the conventional choice for an unknown distribution is the 
Gaussian, there are reasons why it may not be the best in this case.  
As noted in \autoref{sec:Introduction Noise}, the noise is not strictly Gaussian,
nor is it strictly Wiener. As with any unknown noise however, it is necessary 
to make some assumption. If the weighting function ($P(y_k | x_k)$) exactly
matches the measurement error, then the ideal particle filter will result.
Particles with $x_k$'s that repeatedly estimate $y_k$ with large residual 
will quickly have weights near 0. Thus, a weighting function that
exactly matches $P(Y(t) | X(t))$ will easily and correctly throw out invalid
particles.  The cost of choosing an overly broad distribution for this
function is slow convergence.  On the other hand, an overly thin 
distribution will lead to particle deprivation (all particles
being zero-weighted). Three weighting functions were 
tested. In addition to the Gaussian I also tested the Laplace and Cauchy
distributions, both of which have wide tales. 
Wider tailed distributions don't down-weight
particles as fast; and thus converge more slowly. 
The Laplace distribution also has the
benefit of a non-zero slope at the origin; meaning it 
differentiates between particles even near the origin.

After trial and error, for this work I chose the Gaussian with a standard deviation 
of $0.005$. Although using an adaptive weighting function
might be better, in tests this often led to unpredictable results. 
With some work it may be possible to set the standard deviation by
taking a small sample from resting data and using the sample standard deviation.
In this work, however, I set the standard deviation manually at $0.005$,
because it gave the best consistency.

@@

Because the prior is initially represented with significantly more
particles, if for some reason the effective
number of particles stays high, resampling could take a long time to occur.
For this reason, rather than allowing the particle filter to continue on 
with this large number of particles, after 20 seconds have passed the
algorithm forces resampling. The choice of 20 seconds is 
arbitrary, but at the very least it gives the particles time to be weighted.
